---
title: "Instructions"
format: html
includesource: true
execute: 
  eval: false
---

# Patient data

We need some patient data:

```{r}
library(data.table)
library(tidyverse)

if (!fs::file_exists("data.zip")) {
  curl::curl_download(
    "https://github.com/eribul/cs/raw/refs/heads/main/data.zip",
    "data.zip",
    quiet = FALSE
  )
}
```

We used `data.table::fread()` in the previous exercise, which is a very fast and competent reader for CSV files. It is usually preffered, but it can not read data directly from a ZIP archive :-( To avoid unzipping the whole file (we do not need it right now), we can use the old fashion `read.csv` instead:

```{r}
patients <-
  readr::read_csv(unz("data.zip", "data-fixed/patients.csv")) |>
  setDT() |>
  setkey(id)
```

Unfortunately, read.csv does not read dates correctly so this must be fixed:

```{r}
patients[, names(.SD) := lapply(.SD, as.Date), .SDcols = patterns("date")]
```

Sometimes there might be whole columns or rows missing in a dataset We won't have much use for those so let's remove them

```{r}
patients <- janitor::remove_empty(patients, quiet = FALSE)
# A column with only one constant value is also not very interesting
patients <- janitor::remove_constant(patients, quiet = FALSE)
```

# Expectations/validations

What are our expectations of the data? Are those expectations full filled or do we need some additional modifications or perhaps delition of some data? CHe out the [pointblank](https://rstudio.github.io/pointblank/articles/VALID-I.html) package!

```{r}
library(pointblank)
al <- action_levels(warn_at = 1)
```

I'll give you some of my expectations but there should probably be more! What do you expect for `marital`? Would "kulbo" be a valid value here? Is everyone supposed to be either male or female (depends on the context, but so far we might have no reason to believe that any sex would be missing).

1.  Replace all my labels with something more informative below (informative even for a clinician without acces to data!)
2.  Add at least 5 more expectations with infomrative labels

```{r}
patients |>
  create_agent(label = "A very simple example.", actions = al) |>
  col_vals_between(
    where(is.Date),
    as.Date("1900-01-01"),
    as.Date(Sys.Date()),
    na_pass = TRUE,
    label = "WHAT DO WE CHECK HERE? DESCRIBE!"
  ) |>
  col_vals_gte(
    deathdate,
    vars(birthdate),
    na_pass = TRUE,
    label = "ADD SOMETHING!"
  ) |>
  col_vals_regex(
    ssn,
    "{3}-[0-9]{2}-[0-9]{4}$",
    label = "WHAT ON EARTH DOES THIS MEAN?"
  ) |>
  col_is_integer(
    id,
    label = "ADD ..."
  ) |>
  interrogate()
```

## Factor levels

Some columns have just abbreviations for their values. Let's convert those to something more informative. In this case we don't actually know what the different marital status are but me might have an educated guess.

::: callout-warning
## \# Ask for help

To just guess is obviously dangerous in a real world scenario! Lacking documentation, however, is a very real problem in many cases! But always try to confirm your guesses with someone and carefully document what you don't know!
:::

```{r}
patients[, .N, marital]
```

My best guess would be:

```{r}
patients[,
  marital := factor(
    marital,
    levels = c("S", "M", "D", "W"),
    labels = c("Single", "Married", "Divorced", "Widowed")
  )
]
```

Do you agree

The abbriviations used for `gender` actually seems to correspond more to what is expected for `sex`. Do we actually know if this variable concerns `gender` or `sex`? Does it matter in this case? Add factor levels for all variables deemed relevant. When looking at race, you might see that one group is quite small compared to the others. Whould it be sensible to present data in a later stage where individual persons might be able to identify themselves, for example based on a combination of this variable and a rare disease? Do you have the ethical permission to present such data (you obviously had the permission to retrieve the data, but that is not the same thing)? Should certain groups be combined when analyzed?

# Derived variables

How old are these people? We know when they were born, so let's subtract that from todays date and calculate the whole number of times they have circled the sun up until today

```{r}
patients[, age := as.integer((as.Date(Sys.Date()) - birthdate)) %/% 365.241]
patients[, hist(age)]
```

Did we forget something? Maybe we should focus on the living?

```{r}
patients[is.na(deathdate), hist(age)]
```

Hmm, living by the time the data was collected ... but now we look at their age TODAY. Is this right? When was the original data set generated? Let's say we don't now ... (Yes, you should know if this was your own data, but reality is not always perfect). We have data for `payer_transitions`. Since this is US data, such things are important to their health care. We might check when the last transaction was performed and might guess that the data snapshot was taken soon after that.

We only need a single date however, so no need to read the whole file (no problem with this toy data but you might have massive data sets later in life). First, unzip only the relevant file and then use the arrow package and duckdb to perform some calculations on disk.

```{r}
unzip("data.zip", files = "data-fixed/payer_transitions.csv")
lastdate <-
  arrow::open_csv_dataset("data-fixed/payer_transitions.csv") |>
  arrow::to_duckdb() |>
  summarise(lastdate = max(end_date)) |>
  collect() |>
  pluck("lastdate") |>
  as.Date()
```

What date was it? Is it reasonable? Well, probably not! We were not careful enough when chosing the column. The end_date might be in the future for enrolled payers! Use the `start_date` instead! Now present a histogram of the living patients age distribution at the time of the assumed data extract (we have no idea what happend thereafter)!

What would we call those peopple? In a daily conversation we might not need to separate between fisrst and middle names etc.

```{r}
patients[,
  full_name := paste(
    prefix,
    first,
    middle,
    last,
    fifelse(suffix != "", paste0(", ", suffix), "")
  )
]
patients[, full_name]
```

Hm, some people got a trailing white space (can you see why this happened?) Lets remove all such things for all character columns

```{r}
patients[, names(.SD) := lapply(.SD, trimws), .SDcols = is.character]
```

With this new name variable, we might no longer need some of the original columns:

```{r}
patients[, c("prefix", "first", "middle", "last", "suffix", "maiden") := NULL]
```

Data originates from a database. As such, the concept of NA might not have existed in the source (NULL is a slightly different concept). Instead empty strings are used to indicate missing data in some variables. (If using data.table::fread(), this might be handled by setting its `na.strings` argument.)

```{r}
patients[, names(.SD) := lapply(.SD, \(x) na_if(x, "")), .SDcols = is.character]
```

Btw, are empty strings the same as missing data? I think so in this case. But not always! Thnik carefully for each variable!

Do we have any use of the `drivers` column? Mayby if we want to link the data to additional sources. Could be interesting for example to see for how long an individual had his/her license before being involved in a traffic accident for example. We don't have that info now so the only relevant information might be if an individual has a license or not.

```{r}
patients[, driver := !is.na(drivers)]
```

The same probably goes for several other columns. It might be good to deleta them when working with big data (to save space and therefore computational time etc) but in this case we could as well keep some extra columns for now.

The lat and long columns are probably most relevant if we want to calculate distances to other places. Does individuals with a longer/shorter distance to their nearest hospital recieve better/worse care? Is the prevalence of skin cancer more prevalent if you live closer to the sea?

But it might also just be fun to play with visualisations some times (if allowed by the ethical permit etc!). This might also be useful to decide how representative the sample is.

```{r}
leaflet::leaflet(data = patients) |>
  leaflet::addTiles() |>
  leaflet::addMarkers(~lon, ~lat, label = ~full_name)
```

Based on this data, would it be reasonable to make statistical inference for the whole of USA? For the whole world?

There are many ways to work with spatial data (hence data which can be projected onto a map). We will not dvelwe much deeper into this now but you might check out the packages {sf} and {tmap} at some point.

# A simple report

Above, as well as in the previous exercise we have learned a lot of techniques wich might be helpful for data preparation before any statistical analysis. Now, try to think of a simple research qustion which might be answered based on the data at hand. The question does not need to be unique or novel. But make sure it seems at least reasonable.

Try this hypothesis genarator if you are really stuck (but don't just pick the first suggestion if it does not make sense):

```{r}
random_research_question <- function() {
  v <- sample(names(patients))
  adj <- v[3:5] |> paste(collapse = ", ")
  sprintf(
    "Is there an association between %s and %s adjusting for %s",
    v[1],
    v[2],
    adj
  )
}
random_research_question()
```

Your analysis can also be explorative rather than hypothesis driven! Also note that you may use derived variables (such as age instead of birthdate or having a drivers license rather than the license number itself etc).

Use all your skills to make a brief report exploring this question. Present it in a quarto document (use the already prepared `report_cs1.qmd` document). Keep the section headings but modify all other context. Don't change the file name! The document should contain code to load the data, prepare it for analysis, and present the results etc. You can reuse code from the previous exercises or from any source, but if so, copy and paste it into the report and make sure you understand what it is doing (include comments for your future self as well as for other potential readers of your code)!

When you are done, commit all changes to your Git repo and push all changes to the main branch on GitHub. If someone else clone your repo later, they should be able to render your document without access to any other data. Hence, you need to include code for downloading the zip file as well in the beginning of your report! You are allowed to use any R package from CRAN however (without explicitly including `install.packages(_____)`).